\bf Visualising How Non-linear Dimension Reduction Warps Your Data

### 2024-03-19 paper.qmd
Added high-D model lifting with example changed with new hex-binning approach

@@

```{r}
#| eval: false
#| warning: false
#| echo: false
#| label: fig-wkhighD
#| fig-cap: How the 2D model lift into high dimensions? (a) visualize the points and the hexagonal bin centroids related 2nd and 15th hexagons, (b) visualization of the edge connected the 2nd and 15th hexagons (colored in red) in the triangular mesh. A video of tour animation is available at <https://youtu.be/ffWFVK1UTIc>.

>>>>

num_bins_x <- 5
num_bins_y <- 10
hex_size <- 0.19

hb_obj <- hex_binning(data = umap_s_curve_scaled, x = "UMAP1", 
                              y = "UMAP2", num_bins_x = num_bins_x, 
                              num_bins_y = num_bins_y, x_start = NA, y_start = NA, 
                              buffer_x = NA, buffer_y = NA, hex_size = hex_size, col_start = "UMAP")

## Data set with all possible centroids in the hexagonal grid
all_centroids_df <- as.data.frame(do.call(cbind, hb_obj$centroids))

## Generate all coordinates of hexagons
hex_grid <- as.data.frame(do.call(cbind, hb_obj$hex_poly))

## To obtain the standardise counts within hexbins
counts_df <- as.data.frame(do.call(cbind, hb_obj$std_cts))
df_bin_centroids <- extract_hexbin_centroids(centroids_df = all_centroids_df, 
                                             counts_df = counts_df)

UMAP_data_with_hb_id <- as.data.frame(do.call(cbind, hb_obj$data_hb_id))

model_object <- fit_highd_model( training_data = training_data, 
                                 nldr_df_with_id = umap_s_curve_scaled, 
                                 x = "UMAP1", y = "UMAP2", 
                                 num_bins_x = num_bins_x, 
                                 num_bins_y = num_bins_y, 
                                 x_start = NA, y_start = NA, 
                                 buffer_x = NA, buffer_y = NA, 
                                 hex_size = hex_size,
                                 is_rm_lwd_hex = FALSE, 
                                 benchmark_to_rm_lwd_hex = NA, 
                                 col_start_2d = "UMAP", 
                                 col_start_highd = "x")

df_bin_centroids <- model_object$df_bin_centroids
df_bin <- model_object$df_bin

## Triangulate bin centroids
tr1_object <- tri_bin_centroids(df_bin_centroids, x = "c_x", y = "c_y")
tr_from_to_df <- gen_edges(tri_object = tr1_object)

## Compute 2D distances
distance <- cal_2d_dist(tr_coord_df = tr_from_to_df, 
                                start_x = "x_from", start_y = "y_from", 
                                end_x = "x_to", end_y = "y_to", 
                                select_vars = c("from", "to", "distance"))

hb_id_selected <- tr_from_to_df_n |>
  dplyr::filter((from == 1) & (to == 3))

# Filter and label small and long edges
distance_df_small_edges <- distance |>
  dplyr::filter((from != 1) | (to != 3)) |>
  dplyr::mutate(type = "small_edges")

distance_df_long_edges <- distance |>
  dplyr::filter((from == 1) & (to == 3)) |>
  dplyr::mutate(type = "long_edges")

a1 <- ggplot(bin_coord, aes(x = x, y = y)) +
  geom_segment(
    aes(x = x_from, y = y_from, xend = x_to, yend = y_to, colour = type),
    data = tr_from_to_df_coord_with_group) +
  geom_point() +
  #coord_equal() +
  coord_cartesian(xlim =c(-0.5, 1.5), ylim = c(-0.5, 2.7)) +
  geom_point(data = UMAP_data_with_hb_id_selected, aes(x = UMAP1, y = UMAP2, colour = select_point), alpha = 0.5, size = 1) +
  scale_color_manual(values=c("#e6550d", "#000000", "#e31a1c", "#f6e8c3")) +
  geom_point(size = 3, colour = "#33a02c") + 
  #ggtitle("(b)") + 
  xlab(expression(C[x]^{(2)})) + ylab(expression(C[y]^{(2)})) +
  theme_light() +
  #coord_equal() +
  theme(legend.position = "none", plot.title = element_text(size = 5, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()#change legend key width
  )

a2 <- ggplot(data = hex_grid, aes(x = x, y = y)) + 
  geom_polygon(color = "#deebf7", aes(group = hex_poly_id), fill = "#ffffff") +
  geom_segment(
    aes(x = x_from, y = y_from, xend = x_to, yend = y_to), colour = "#f6e8c3",
    data = tr_from_to_df_coord_with_group
  ) +
  geom_point(data = UMAP_data_with_hb_id_selected, aes(x = UMAP1, y = UMAP2, colour = select_point), size = 1, alpha = 0.5) +
  scale_color_manual(values=c("#e6550d", "#000000")) +
  geom_point(data = df_bin_centroids |> dplyr::filter(hexID %in% c(hb_id_selected$from_hb_id, hb_id_selected$to_hb_id)), aes(x = c_x, y = c_y), colour = "#33a02c", size = 3) +
  geom_point(data = UMAP_data_with_hb_id_selected_no, aes(x = UMAP1, y = UMAP2), colour = "#6a3d9a", size = 1, alpha = 0.5) +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "C") +
  coord_equal() +
  theme_light() +
  theme(legend.position = "none", plot.title = element_text(size = 5, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()#change legend key width
  )

lg_high <- langevitour::langevitour(df_exe[1:(length(df_exe)-1)], group = df_exe$type, pointSize = append(rep(4, NROW(df_b)), rep(2, NROW(df1) + NROW(df2) + NROW(df3))), levelColors = c("#6a3d9a", "#e6550d", "#000000", "#33a02c"), lineFrom = 1, lineTo = 2, lineColors = "#e31a1c")

```{r}
#| warning: false
#| echo: false
#| label: fig-wkhighD
#| fig-cap: How the 2D model lift into high dimensions? (a) visualize the points and the hexagonal bin centroids related $7^{th}$ and $12^{th}$ hexagons, (b) visualization of the edge connected the $7^{th}$ and $12^{th}$ hexagons (colored in red) in the triangular mesh. A video of tour animation is available at <https://youtu.be/hxU91xNTJL0>.

a2 + a1 +
  plot_annotation(tag_levels = 'a') +
  plot_layout(ncol = 3) &
  theme(legend.position='none', plot.tag = element_text(size = 8))
```

@@

```{r}
#| eval: false
#| warning: false
#| echo: false
#| label: fig-wkhighD1
#| fig-cap: How the 2D model lift into high dimensions? (a) visualize the points and the hexagonal bin centroids related 2nd and 15th hexagons, (b) visualization of the edge connected the 2nd and 15th hexagons (colored in red) in the triangular mesh. A video of tour animation is available at <https://youtu.be/ffWFVK1UTIc>.

### 2024-03-19 paper.tex
Added high-D model lifting with example changed with new hex-binning approach

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, breakable, sharp corners, borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

As described in Step 4 of Section~\ref{sec-construct2d}, during the
triangulation process in 2D space, vertices are identified to form
edges. With the knowledge of the high-D mappings for the 2D hexagonal
bins, the vertices connected in 2D are also connected in high-D (see
video linked in \textbf{?@fig-wkhighD}).

@@

\includegraphics{paper_files/figure-pdf/unnamed-chunk-85-1.pdf}

>>>>

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, breakable, borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, boxrule=0pt, frame hidden]}{\end{tcolorbox}}\fi

@@

As described in Step 4 of Section~\ref{sec-construct2d}, during the
triangulation process in 2D space, vertices are identified to form
edges. With the knowledge of the high-D mappings for the 2D hexagonal
bins, the vertices connected in 2D are also connected in high-D (see
video linked in Figure~\ref{fig-wkhighD}).

{\centering \includegraphics{paper_files/figure-pdf/fig-wkhighD-1.pdf}

\caption{\label{fig-wkhighD}How the 2D model lift into high dimensions?
(a) visualize the points and the hexagonal bin centroids related
\(7^{th}\) and \(12^{th}\) hexagons, (b) visualization of the edge
connected the \(7^{th}\) and \(12^{th}\) hexagons (colored in red) in
the triangular mesh. A video of tour animation is available at
\url{https://youtu.be/hxU91xNTJL0}.}

@@

\includegraphics{paper_files/figure-pdf/unnamed-chunk-87-1.pdf}
